---
  "Introduction to several commonly used functions in R"
--

## with()
The with( ) function applys an expression to a dataset. It is similar to 
DATA= in SAS.

```{r}
# with(data, expression)
# example applying a t-test to a data frame mydata 
with(mydata, t.test(y ~ group))

```

Please look at other examples [here](http://stackoverflow.com/questions/21827572/what-is-the-difference-between-with-and-within-in-r) and [here](http://www.magesblog.com/2012/01/say-it-in-r-with-by-apply-and-friends.html).

## by()
The by( ) function applys a function to each level of a factor or factors. 
It is similar to BY processing in SAS.

```{r}
# by(data, factorlist, function)
# example obtain variable means separately for
# each level of byvar in data frame mydata 
by(mydata, mydatat$byvar, function(x) mean(x))
```


## do.call()
do.call calls a function with a list of arguments, lapply applies a function to each element of the list

```{r}
do.call(sum, list(c(1,2,4,1,2), na.rm = TRUE))
#10
lapply(c(1,2,4,1,2), function(x) x + 1)
#2
#3
#5
#2
#3
do.call("+",list(4,5))
#9
```
