---
title: "Introduction to dplyr for Faster Data Manipulation in R"
output: html_document
---

Note: There is a 40-minute [video tutorial](https://www.youtube.com/watch?v=jWjqLW-u3hc) on YouTube that walks through this document in detail.

## Why do I use dplyr?

* Great for data exploration and transformation
* Intuitive to write and easy to read, especially when using the "chaining" syntax (covered below)
* Fast on data frames


## dplyr functionality

* Five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarise` (plus `group_by`)
* Can work with data stored in databases and [data tables](http://datatable.r-forge.r-project.org/)
* Joins: inner join, left join, semi-join, anti-join (not covered below)
* Window functions for calculating ranking, offsets, and more
* [Better than plyr](http://blog.rstudio.org/2014/01/17/introducing-dplyr/) if you're only working with data frames (though it doesn't yet duplicate all of the plyr functionality)
* Examples below are based upon the [latest release](https://github.com/hadley/dplyr/releases), version 0.2 (released May 2014)


## Loading dplyr and an example dataset

* dplyr will mask a few base functions
* If you also use plyr, load plyr first
* hflights is flights departing from two Houston airports in 2011

```{r}
# load packages
suppressMessages(library(dplyr))
library(hflights)

# explore data
data(hflights)
head(hflights)
```

* `tbl_df` creates a "local data frame"
* Local data frame is simply a wrapper for a data frame that prints nicely

```{r}
# convert to local data frame

flights <- tbl_df(hflights)

# printing only shows 10 rows and as many columns as can fit on your screen
flights
```

```{r results='hide'}
# you can specify that you want to see more rows
print(flights, n=20)

# convert to a normal data frame to see all of the columns
data.frame(head(flights))
```


## filter: Keep rows matching criteria

* Base R approach to filtering forces you to repeat the data frame's name
* dplyr approach is simpler to write and read
* Command structure (for all dplyr verbs):
    * first argument is a data frame
    * return value is a data frame
    * nothing is modified in place
* Note: dplyr generally does not preserve row names

```{r results='hide'}
# base R approach to view all flights on January 1
flights[flights$Month==1 & flights$DayofMonth==1, ]
#another method
subset(flights,flights$Month==1 & flights$DayofMonth==1)
subset(flights,Month==1 & DayofMonth==1)
```

```{r}
# dplyr approach
# note: you can use comma or ampersand to represent AND condition
filter(flights, Month==1, DayofMonth==1)

# use pipe for OR condition
filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA")
```

```{r results='hide'}
# you can also use %in% operator
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```


## select: Pick columns by name

* Base R approach is awkward to type and to read
* dplyr approach uses similar syntax to `filter`
* Like a SELECT in SQL

```{r results='hide'}
# base R approach to select DepTime, ArrTime, and FlightNum columns
flights[, c("DepTime", "ArrTime", "FlightNum")]
```

```{r}
# dplyr approach
select(flights, DepTime, ArrTime, FlightNum)

# use colon to select multiple contiguous columns, and use `contains` to match columns by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))
```


## "Chaining" or "Pipelining"

* Usual way to perform multiple operations in one line is by nesting
* Can write commands in a natural order by using the `%>%` infix operator (which can be pronounced as "then")

```{r results='hide'}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)
```

```{r}
# chaining method
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    filter(DepDelay > 60)
```

* Chaining increases readability significantly when there are many commands
* Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
* Can be used to replace nesting in R commands outside of dplyr

```{r results='hide'}
# create two vectors and calculate Euclidian distance between them
x1 <- 1:5; x2 <- 2:6
sqrt(sum((x1-x2)^2))
```

```{r}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```


## arrange: Reorder rows

```{r results='hide'}
# base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]
```

```{r}
# dplyr approach
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay)
```

```{r results='hide'}
# use `desc` for descending
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```


## mutate: Add new variables

* Create new variables that are functions of existing variables

```{r results='hide'}
# base R approach to create a new variable Speed (in mph)
flights$Speed <- flights$Distance / flights$AirTime*60
flights[, c("Distance", "AirTime", "Speed")]
```

```{r}
# dplyr approach (prints the new variable but does not store it)
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance/AirTime*60)

# store the new variable
flights <- flights %>% mutate(Speed = Distance/AirTime*60)
```


## summarise: Reduce variables to values

* Primarily useful with data that has been grouped by one or more variables
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r results='hide'}
# base R approaches to calculate the average arrival delay to each destination
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm=TRUE)))
head(aggregate(ArrDelay ~ Dest, flights, mean))
```

```{r}
# dplyr approach: create a table grouped by Dest, and then summarise each group by taking the mean of ArrDelay
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm=TRUE))
```

* `summarise_each` allows you to apply the same summary function to multiple columns at once
* Note: `mutate_each` is also available

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(mean), Cancelled, Diverted)

# for each carrier, calculate the minimum and maximum arrival and departure delays
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("Delay"))
```

* Helper function `n()` counts the number of rows in a group
* Helper function `n_distinct(vector)` counts the number of unique items in that vector

```{r}
# for each day of the year, count the total number of flights and sort in descending order
flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))

# rewrite more simply with the `tally` function
flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)

# for each destination, count the total number of flights and the number of distinct planes that flew there
flights %>%
    group_by(Dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(TailNum))
```

* Grouping can sometimes be useful without summarising

```{r}
# for each destination, show the number of cancelled and not cancelled flights
flights %>%
    group_by(Dest) %>%
    select(Cancelled) %>%
    table() %>%
    head()
```


## Window Functions

* Aggregation function (like `mean`) takes n inputs and returns 1 value
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) takes n inputs and returns n values
    * Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).

```{r results='hide'}
# for each carrier, calculate which two days of the year they had their longest departure delays
# note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    filter(min_rank(desc(DepDelay)) <= 2) %>%
    arrange(UniqueCarrier, desc(DepDelay))
```

```{r}
# rewrite more simply with the `top_n` function
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    top_n(2) %>%
    arrange(UniqueCarrier, desc(DepDelay))

# for each month, calculate the number of flights and the change from the previous month
flights %>%
    group_by(Month) %>%
    summarise(flight_count = n()) %>%
    mutate(change = flight_count - lag(flight_count))

# rewrite more simply with the `tally` function
flights %>%
    group_by(Month) %>%
    tally() %>%
    mutate(change = n - lag(n))
```


## Other Useful Convenience Functions

```{r}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)

# randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace=TRUE)

# base R approach to view the structure of an object
str(flights)

# dplyr approach: better formatting, and adapts to your screen width
glimpse(flights)
```


## Connecting to Databases

* dplyr can connect to a database as if the data was loaded into a data frame
* Use the same syntax for local data frames and databases
* Only generates SELECT statements
* Currently supports SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB
* Example below is based upon an SQLite database containing the hflights data
    * Instructions for creating this database are in the [databases vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)

```{r}
# connect to an SQLite database containing the hflights data
my_db <- src_sqlite("my_db.sqlite3")

# connect to the "hflights" table in that database
flights_tbl <- tbl(my_db, "hflights")

# example query with our data frame
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))

# identical query using the database
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```

* You can write the SQL commands yourself
* dplyr can tell you the SQL it plans to run and the query execution plan

```{r}
# send SQL commands to the database
tbl(my_db, sql("SELECT * FROM hflights LIMIT 100"))

# ask dplyr for the SQL commands
flights_tbl %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay)) %>%
    explain()
```


## Resources

* [Official dplyr reference manual and vignettes on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html): vignettes are well-written and cover many aspects of dplyr
* [July 2014 webinar about dplyr (and ggvis) by Hadley Wickham](http://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html) and related [slides/code](https://github.com/rstudio/webinars/tree/master/2014-01): mostly conceptual, with a bit of code
* [dplyr tutorial by Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) at the [useR! 2014 conference](http://user2014.stat.ucla.edu/): excellent, in-depth tutorial with lots of example code (Dropbox link includes slides, code files, and data files)
* [dplyr GitHub repo](https://github.com/hadley/dplyr) and [list of releases](https://github.com/hadley/dplyr/releases)


< END OF DOCUMENT >